---
title: "Racial Disparities among Re-Entry Outcomes:"
subtitle: "Example Research Problem and Analysis Solutions"
author: "Mari Roberts"
date: "4/2/2021"
output: pdf_document
---

Client Generated Research Question: But before I build policy and practice, I need some facts regarding what’s going on. Do people of color experience a higher number re-entry challenges than whites, and does this impact their ability to remain out of prison?

**Data Details:** Includes short surveys of 1,020 individuals, recently released from state prison in Maricopa County, Arizona and tracked over a 2-year time period. There are three waves of data: wave 1 takes places about 1 month before the individual is released from prison, wave 2 takes place about 1-year post-release, and wave 3 takes place about 2 years post-release. Personal characteristics that are time-invariant assessed prior to release from prison. Time-varying characteristics are assessed at 1- and 2-years post-release from prison. Recidivism is defined as re-incarceration or returning to prison and verified using administrative data (i.e., ADOC confirms the respondent was re-incarcerated). While all individuals were surveyed regardless of whether or not they were incarcerated at either waves 2 or 3, some time-varying characteristics will be not applicable if the individual has been re-incarcerated.

# Data Cleaning

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# load necessary packages
requiredPackages = c('dplyr',
                     'readr',
                     'readxl',
                     'haven',
                     'ggplot2',
                     'tidyverse')
# only downloads packages if needed
for(p in requiredPackages){
  if(!require(p,character.only = TRUE)) install.packages(p)
  library(p,character.only = TRUE)
}

# custom theme
theme_csgjc <- theme(axis.ticks = element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_text(vjust = 8, margin = margin(t = 6),size=11,face="bold"),
                     #axis.title.y = element_blank(),
                     #axis.title.x = element_blank(),
                     panel.border = element_blank(), 
                     panel.grid.major.x = element_blank(), 
                     panel.grid.minor.x = element_blank(), 
                     panel.grid.major.y = element_blank(), 
                     panel.grid.minor.y = element_blank(),
                     legend.title = element_blank(),
                     legend.position = "right",
                     plot.title = element_text(hjust = 0.5,size = 12, face = "bold"),
                     plot.subtitle = element_text(hjust = 0.5, size = 15),
                     plot.margin = margin(0, 0, 0, 0, "cm"))

# import data
reentry.dta <- read_dta("4_2_2021/reentry_data.dta")

# duplicate for cleaning
reentry <- reentry.dta

# view data types
str(reentry)

# change to num variables
reentry$age <- as.numeric(reentry$age)
reentry$edu <- as.numeric(reentry$edu)
reentry$lsir <- as.numeric(reentry$lsir)
reentry$msf12 <- as.numeric(reentry$msf12)
reentry$sf12 <- as.numeric(reentry$sf12)
reentry$instrumentalss <- as.numeric(reentry$instrumentalss)
reentry$emotionalss <- as.numeric(reentry$emotionalss)

# change to factor variables
names <- c("id","wave","race","white","black","asian","other","latinx","female","prioremp",
           "partner","kids","pchronic","mchronic","disab","courtfees","recid","o10","employed",     
           "empft","parole","ins","moneysource","nhd","aodtx","housingtype")
reentry[,names] <- lapply(reentry[,names], factor)   

# change race factor levels
# reentry <- reentry %>% mutate(race2 = factor(race, levels = c("white", "black", "asian", "other", "latinx")))
reentry$race <- case_when(reentry$race == 1 ~ "white",
                               reentry$race == 2 ~ "black",
                               reentry$race == 3 ~ "asian",
                               reentry$race == 4 ~ "other",
                               reentry$race == 5 ~ "latinx")

# count unique people = 1020
length(unique(reentry$id))

# race data is missing in wave 1
# add race based on other wave info
reentry <- reentry %>% 
  group_by(id) %>% 
  fill(race, .direction = "up")
```

# Descriptive Statistics

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# view data
table(reentry$employed,reentry$race)
table(reentry$housingtype, reentry$race)
table(reentry$nhd, reentry$race)

# create a concerned variable based on mental health 
reentry <- reentry %>% mutate(concerned = ifelse(msf12 > 2 | msf12 < -2,1,0))

# chi-square tests
chisq.test(table(reentry$race, reentry$employed))
chisq.test(table(reentry$race, reentry$housingtype))
chisq.test(table(reentry$race, reentry$nhd))
chisq.test(table(reentry$race, reentry$msf12))
chisq.test(table(reentry$race, reentry$concerned))
```

# Data Visualizations

## Race and Mental Health

```{r}
# historgram of mental health
ggplot(data=reentry, aes(reentry$msf12)) + 
  geom_histogram(color="black", fill="white") + theme_bw() + theme_csgjc
```

## Race and Housing
```{r}
# race and housing
df <- reentry %>% filter(wave==2) %>% group_by(race, housingtype) %>% summarise(total = n())
df <- df %>% filter(race == "white" | race == "black" | 
                      race == "asian" | race == "other" | 
                      race == "latinx")

# grouped barplot
ggplot(df, aes(fill=race, y=total, x=housingtype)) + 
    geom_bar(position="dodge", stat="identity") + theme_bw() + theme_csgjc +   
    geom_text(aes(label = scales::comma(round(total), accuracy=1)), 
              position = position_dodge2(width = 0.9, preserve = "single"),
              vjust=-.5, hjust=.5,color="black", size=3.5, width = 0.65,fontface = "bold")
```
